A Survey of Real-Time Strategy Game AI\\ Research and Competition in StarCraft

### 2013-03-08 survey.tex
Reviewing chapters I and II

Many  challenging AI  problems can  be found  in  dynamic environments
where  one does  not have  a  perfect, complete  information of  their
current  state, as well  as their  evolution.  Road  traffic, finance,
weather  forecasts are  examples of  huge, complex,  real-life dynamic
environments. RTS games can be seen as a simplification of a real-life
environment, where there  exist simpler rules in a  finite and smaller
world, but are complex enough  to bring very difficult and interesting
problems  to  the scientific  community,  like  decision making  under
uncertainty  or  real-time adversarial planning.   Finding  efficient  techniques
tackling  these problems  on  RTS  games can  thus  profit broader  AI
domains, but  have also concrete  and direct applications in  the huge
industry of  video games,  generating worldwide two  and a  half times
more revenue than the Film industry in 2011.

This paper, aiming to provide a one-stop guide on what is the state of
the art in RTS AI, and with a particular emphasis on the work done in StarCraft, is organized as follows: Section~\ref{sec:rts}
introduces real-time strategy games, in particular the game StarCraft,
and  the main  AI challenges  in RTS  games.  Section~\ref{sec:review}
reviews  the existing  work on  RTS games  tackling  these challenges.
Section~\ref{sec:bot} shows  the current state  of the art  AI players
for  StarCraft, presenting  seven AI  players participating  to annual
StarCraft    AI    competitions     for    a    couple    of    years.
Section~\ref{sec:competition}  presents these two  annual competitions
held  at the AIIDE  and CIG  conferences.  Section~\ref{sec:questions}
compiles open questions  in RTS game AI which must  be tackled to make
bots reaching a high human-level play.
% skip "empirical evidence..." for the moment
Finally, the paper concludes on discussions and perspectives.

@@

% {\color{red} The term bot is introduced in II.B.2 Learning, when we talk about
% bots for the first time in the paper.}
% {\color{red} make sure to introduce the term ``bot''}

@@

\begin{itemize}
\item They are {\em simultaneous move} games, where more than one player can issue actions at the same time. Additionally, these actions are {\em durative}, i.e. actions are not instantaneous, but take some amount of time time to complete.
\item RTS games are ``real-time'', which actually means is that each player has a very small amount of time to decide the next move. Compared to Chess, where players have several minutes to decide the next action, in StarCraft, the game executes at 24 iterations per second, which means that players only have 55ms to decide a move, before the game state changes.
\item Most RTS games are partially observable: players can only see the part of the map that has been explored.
\item Most RTS games are non-deterministic.
\item And finally, the complexity of these games, both in terms of state space size and in terms of number of actions available at each decision cycle is very large. For example, the state space of Chess is typically estimated to be around $10^{50}$, heads up no-limit texas holdem poker around $10^{80}$, and Go around $10^{170}$. In comparison, the state space of StarCraft in a typical map is estimated to be many orders of magnitude larger than any of those, as discussed in the next section.
\end{itemize}

For those reasons, standard techniques used for playing classic board games, such as alpha-beta search, cannot be directly applied to solve RTS games without the definition of some level of abstraction, or some other simplification. Interestingly enough, humans seem to be able to deal with the complexity of RTS games, and are still vastly superior to computers in these type of games \cite{burochurchill2012aimagazine}. For those reasons, a large spectrum of techniques have been attempted to deal with this domain, as we will describe below. The remainder of this section is devoted to describe StarCraft as a research testbed, and on detailing the open challenges in RTS game AI.

@@

{\em StarCraft: Brood War} is an immensely popular RTS game released in 1998 by Blizzard Entertainment. StarCraft is set in a science-fiction based world where the player must choose one of the three races: Terran, Protoss or Zerg. One of the most remarkable aspects of StarCraft is that the three races in StarCraft are extremely well balanced:

@@

Figure~\ref{fig:StarCraft} shows a screenshot of StarCraft showing a player playing Terrans. In order to win a StarCraft game, players must first gather resources (minerals and gas), which they can spend in creating additional buildings and units. As resources become available, players need to allocate them for creating more buildings (which reinforce the economy, and allow players to create units or unlock stronger units) and train attack units. Units must be distributed to accomplish different tasks such as reconnaissance, defense and attack. While performing all of those tasks, players also need to strategically understand the geometry of the map at hand, in order to decide where to place new buildings (concentrate in a single area, or expand to different areas) or where to set defensive outposts. Finally, when offensive units of two players meet, each player must quickly maneuver each of the units in order to fight a battle, which requires quick and reactive control of each of the units in order to successfully fight the battle.

\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/starcraft1.png}
    \caption{A screenshot of StarCraft% {\color{red} is this a good screenshot? can we actually have a screenshot? or is this copyrighted? Do we even want a screenshot?}.
    .}
    \label{fig:StarCraft}
\end{figure}

A typical StarCraft map is defined as a rectangular grid,
typically  ranging between 64x64  to 128x128  in size  (although the
resolution  at which  units  move  is finer),  and  each player  can
control up  to 200 units (not including  buildings).  Moreover, each
different race  contains between 30  to 35 different types  of units
and  buildings,  each  one  with  a significant  number  of  special
abilities. All these factors  together make StarCraft a significant
challenge,   in   which   humans    are   still   much   better   than
computers. Indeed, best StarCraft AIs are ranked between $D$ and $D^-$
on the iCCup\footnote{http://www.iccup.com/StarCraft/} ranking system,
where average  amateur players are  ranked between $C^+$ and  $B$. For
comparison, StarCraft professional  players are usually ranked between
$A^-$ and $A^+$.

@@

\subsubsection{Planning}
As mentioned above, the size of the state space in RTS games is vastly larger than that of traditional board-games such as Chess or Go. Additionally, the number of actions that can be executed at given instant of time is also much larger. Thus, standard adversarial planning approaches, such as game tree search are not directly applicable. As we elaborate later, planning in RTS games can be seen as having multiple levels of abstraction: at a higher level, players need long-term planning capabilities, in order to develop a strong economy in the game; at a low level, individual units need to be moved in coordination to fight battles taking into account the terrain and the opponent. Techniques that can address this large planning problems by either sampling, or hierarchical decomposition do not yet exist.

\subsubsection{Learning}
Given the difficulties in playing RTS games by directly using adversarial planning techniques, many research groups have turned attention to learning techniques. We can distinguish three types of learning problems in RTS games:
\begin{itemize}
\item {\em Prior learning}: how can we exploit available data, such as existing replays, or information about specific maps for learning appropriate strategies before hand? A significant amount of work has gone in this direction.%, such as \cite{WeberCIG09,SynnaeveCIG11,SynnaeveAIIDE11,OntanonMSR10}.
\item {\em In-game learning}: how can bots, \textit{i.e.,} AI players, can deploy online learning techniques that allow them to improve their game play while playing a game? these techniques might include reinforcement learning techniques, but also opponent modeling. The main problem again is the fact that the state space is too large.
\item {\em Inter-game learning}: what can be learned from one game that can be used to increase the chances of victory in the next game? Some work has used simple game-theoretical solutions to select amongst a pool of predefined strategies, but the general problem remains unanswered.
\end{itemize}

@@

% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
% [3], and 53.125% according  to BWAPI website. Who's right? Should we
% add this in the paper?]
%In StarCraft, it is always an advantage to  have its units  on high ground  whereas enemy is  on low ground, since units  on low ground have no vision  on high ground, and firing from low  ground have only 70\% of chance to  hit units on high ground.

@@

\subsubsection{Domain Knowledge Exploitation}
In traditional board-games such as Chess, researchers have exploited the large amounts of existing domain knowledge to create good evaluation functions to be used by alpha-beta search algorithms, extensive opening books, or end-game tables. In the case of RTS games, it is still unclear the ways in which the significantly large amount of domain knowledge (in the forms or strategy guides, replays, etc.) can be exploited by bots. Most work in this area has focused on two main directions: on the one hand, researchers are finding ways in which to hard-code existing strategies into bots, so that bots only need to decide which strategies to deploy, instead of having to solve the complete problem of deciding which actions to execute by each individual unit at each time step. One the other hand, large datasets of replays have been created \cite{synnaeve2012dataset}, from where strategies, trends or plans have been tried to learn. However, StarCraft games are quite complex, and how to automatically learn from such datasets is still an open problem.

\subsubsection{Task Decomposition}
For all the previous reasons, most existing approaches to play games like StarCraft work by decomposing the problem of playing a RTS game into a collection of smaller problems, to be solved independently. Specifically, a common subdivision is:
\begin{itemize}
\item {\em Strategy}: often called {\em macro-management}, corresponds to the high-level decision making process.  This is
  the  highest level of  abstraction for  the game  comprehension. RTS
  games   are   all   about   finding   an   efficient   strategy   or
  counter-strategy  against a  given opponent.  It concerns  the whole
  set of  units and buildings  one owns. %If  we may allow  ourselves a  military analogy, strategy is orders given by generals at the HQ. 
\item  {\em Tactics}: are  the implementation  of the  current  strategy. It
  implies  army and  building positioning,  movements, timing,  and so
  on.  Tactics concerns a  group of  units. %They  are orders  given by officers on the battlefield.
\item  {\em Reactive  control}: often called, {\em micro-management} is  the  implementation  of  tactics.  This
  consists   in  moving,   targeting,  firing,   fleeing,  hit-and-run
  techniques (aka ``kiting'')  during battle. Reactive control focuses
  on a specific unit. %It is soldiers applying officers' orders. 
%\end{itemize}
%
%\noindent
%Both     strategy      and     tactics     are      usually     called
%\textit{macro-management},  whereas reactive  control  is also  called
%\textit{micro-management}. In addition of the three tasks shown above,
%one could also consider these two tasks that an AI has to deal with:
%\begin{itemize}
\item {\em Terrain analysis}: consists  in the analysis of regions composing
  the  map:  choke-points,  minerals  and  gas emplacements,  low  and  high
  walkable grounds, islands, etc.
\item {\em Intelligence gathering}:  corresponds to information collected about
  the  opponent. Because of  the fog-of-war,  players must
  regularly send scouts to localize and spy enemy bases.%, and must see or  infer what buildings and technology the enemy has developed in order  to adapt its strategy.
\end{itemize}

@@

On the other hand, bots such as NOVA or BroodwarBotQ (BBQ) only use a hierarchical organization for {\em micro-management} (controlling the attack units), but use a decentralized organization for the rest of the bot. In Nova and BBQ, there is a collection of modules that control different aspects of the game (workers, production, construction, etc.). These modules can all send actions directly to StarCraft. In Nova those modules coordinate mostly through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator.

>>>>

Many  challenging AI  problems can  be found  in  dynamic environments
where  one does  not have  a  perfect, complete  information of  their
current  state, as well  as their  evolution.  Road  traffic, finance,
weather  forecasts are  examples of  huge, complex,  real-life dynamic
environments. RTS games can be seen as a simplification of a real-life
environment, where there  exist simpler rules in a  finite and smaller
world, but are complex enough  to bring very difficult and interesting
problems  to  the scientific  community,  like  decision making  under
uncertainty  or  real-time adversarial planning.   Finding  efficient  techniques
to address these problems  on  RTS  games can  thus  profit broader  AI
domains, but  have also concrete  and direct applications in  the huge
industry of  video games.
%,  generating worldwide two  and a  half times more revenue than the Film industry in 2011.
% This statement is not fair, the reference I found it only compare all the game industry revenue (including hardware sales) vs theatrical film industry revenue (i.e. not DVD sales, ...)

This paper, aiming to provide a one-stop guide on what is the state of
the art in RTS AI, and with a particular emphasis on the work done in StarCraft. 
It is organized as follows: Section~\ref{sec:rts}
introduces real-time strategy games, in particular the game StarCraft,
and  the main  AI challenges  in RTS  games.  Section~\ref{sec:review}
reviews  the existing  work on  RTS games  tackling  these challenges.
Section~\ref{sec:bot} shows  the current state  of the art of bots, i.e. a computer-controlled players, 
for  StarCraft, presenting  seven bots participating  to annual
StarCraft    AI    competitions     for    a    couple    of    years.
Section~\ref{sec:competition}  presents these two  annual competitions
held  at the AIIDE  and CIG  conferences and a StarCraft bot game ladder\footnote{An extended tournament, which can potentially go on indefinitely.}.  Section~\ref{sec:questions}
compiles open questions  in RTS game AI which must  be tackled to make
bots reaching a high human-level play.
% skip "empirical evidence..." for the moment
Finally, the paper concludes on discussions and perspectives.

@@

@@

\begin{itemize}
\item They are {\em simultaneous move} games, where more than one player can issue actions at the same time. Additionally, these actions are {\em durative}, i.e. actions are not instantaneous, but take some amount of time time to complete.
\item RTS games are ``real-time'', which actually means is that each player has a very small amount of time to decide the next move. Compared to Chess, where players have several minutes to decide the next action, in StarCraft, the game executes at 24 iterations per second, which means that players only have 55ms to decide a move, before the game state changes.
\item Most RTS games are partially observable: players can only see the part of the map that has been explored.
\item Most RTS games are non-deterministic. Some actions have a chance of success.
\item And finally, the complexity of these games, both in terms of state space size and in terms of number of actions available at each decision cycle is very large. For example, the state space of Chess is typically estimated to be around $10^{50}$, heads up no-limit texas holdem poker around $10^{80}$, and Go around $10^{170}$. In comparison, the state space of StarCraft in a typical map is estimated to be many orders of magnitude larger than any of those, as discussed in the next section.
\end{itemize}

For those reasons, standard techniques used for playing classic board games, such as game tree search, cannot be directly applied to solve RTS games without the definition of some level of abstraction, or some other simplification. Interestingly enough, humans seem to be able to deal with the complexity of RTS games, and are still vastly superior to computers in these type of games \cite{burochurchill2012aimagazine}. For those reasons, a large spectrum of techniques have been attempted to deal with this domain, as we will describe below. The remainder of this section is devoted to describe StarCraft as a research testbed, and on detailing the open challenges in RTS game AI.

@@

{\em StarCraft: Brood War} is an immensely popular RTS game released in 1998 by Blizzard Entertainment. StarCraft is set in a science-fiction based world where the player must choose one of the three races: Terran, Protoss or Zerg. One of the most remarkable aspects of StarCraft is that the three races are extremely well balanced:

@@

Figure~\ref{fig:StarCraft} shows a screenshot of StarCraft showing a player playing Terrans. In order to win a StarCraft game, players must first gather resources (minerals and Vespene gas). As resources become available, players need to allocate them for creating more buildings (which reinforce the economy, and allow players to create units or unlock stronger units), research new technologies (in order to use new unit abilities or improve the units) and train attack units. Units must be distributed to accomplish different tasks such as reconnaissance, defense and attack. While performing all of those tasks, players also need to strategically understand the geometry of the map at hand, in order to decide where to place new buildings (concentrate in a single area, or expand to different areas) or where to set defensive outposts. Finally, when offensive units of two players meet, each player must quickly maneuver each of the units in order to fight a battle, which requires quick and reactive control of each of the units in order to successfully fight the battle.

\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{figures/starcraft1.png}
    \caption{A screenshot of \emph{StarCraft: Brood War}% {\color{red} is this a good screenshot? can we actually have a screenshot? or is this copyrighted? Do we even want a screenshot?}.
    .}
    \label{fig:StarCraft}
\end{figure}

A typical StarCraft map is defined as a rectangular grid, where the $width \times height$ of the map is measured in the number of $32 \times 32$ square of pixels, a.k.a build tiles. Although the
resolution  at which  units  move  is in squares of $8 \times 8$ pixels, a.k.a walk tiles. The standard dimensions are 64, 96, 128, 192 and 256. Each player  can
control up  to 200 units (not including  buildings).  Moreover, each
different race  contains between 30  to 35 different types  of units
and  buildings,  each  one  with  a significant  number  of  special
abilities. All these factors  together make StarCraft a significant
challenge,   in   which   humans    are   still   much   better   than
computers. For instance, in the game ladder iCCup\footnote{http://www.iccup.com/StarCraft/} where users are ranked by their current point totals: $E$ being the lowest possible rank, and $A^+$ and $Olympic$ being the second highest and highest ranks, respectively. Best StarCraft AI bots are ranked between $D$ and $D^−$, where average  amateur players are  ranked between $C^+$ and  $B$. For comparison, StarCraft professional  players are usually ranked between
$A^-$ and $A^+$.

@@

\subsubsection{Planning}
As mentioned above, the size of the state space in RTS games is vastly larger than that of traditional board-games such as Chess or Go. Additionally, the number of actions that can be executed at a given instant of time is also much larger. Thus, standard adversarial planning approaches, such as game tree search are not directly applicable. As we elaborate later, planning in RTS games can be seen as having multiple levels of abstraction: at a higher level, players need long-term planning capabilities, in order to develop a strong economy in the game; at a low level, individual units need to be moved in coordination to fight battles taking into account the terrain and the opponent. Techniques that can address this large planning problems by either sampling, or hierarchical decomposition do not yet exist.

\subsubsection{Learning}
Given the difficulties in playing RTS games by directly using adversarial planning techniques, many research groups have turned attention to learning techniques. We can distinguish three types of learning problems in RTS games:
\begin{itemize}
\item {\em Prior learning}: How can we exploit available data, such as existing replays, or information about specific maps for learning appropriate strategies before hand? A significant amount of work has gone in this direction.%, such as \cite{WeberCIG09,SynnaeveCIG11,SynnaeveAIIDE11,OntanonMSR10}.
\item {\em In-game learning}: How can bots deploy online learning techniques that allow them to improve their game play while playing a game? These techniques might include reinforcement learning techniques, but also opponent modeling. The main problem again is the fact that the state space is too large and the fact that RTS games are partially observable.
\item {\em Inter-game learning}: What can be learned from one game that can be used to increase the chances of victory in the next game? Some work has used simple game-theoretical solutions to select amongst a pool of predefined strategies, but the general problem remains unanswered.
\end{itemize}

@@

% [Flo:  70% to hit high  ground units according to  Ben Weber's paper
% [3], and 53.125% according  to BWAPI website. Who's right? Should we
% add this in the paper?]
% [Alberto: I guess Weber's number comes from StarCraft walkthroughs like this: http://www.supercheats.com/pc/walkthroughs/starcraftbroodwar-walkthrough01.txt, there is also a chance to hit in even grounds (like 98%). But since we cannot see the original code all of these numbers are approximations. In any case we can keep explaining the concept without numbers ;)]
For instance, in StarCraft, it is always an advantage to  have its units  on high ground  whereas enemy is  on low ground, since units  on low ground have no vision  on high ground, and firing from low ground have a chance to  hit units on high ground.

@@

\subsubsection{Domain Knowledge Exploitation}
In traditional board-games such as Chess, researchers have exploited the large amounts of existing domain knowledge to create good evaluation functions to be used by alpha-beta search algorithms, extensive opening books, or end-game tables. In the case of RTS games, it is still unclear the ways in which the significantly large amount of domain knowledge (in the forms or strategy guides, replays, etc.) can be exploited by bots. Most work in this area has focused on two main directions: on the one hand, researchers are finding ways in which to hard-code existing strategies into bots, so that bots only need to decide which strategies to deploy, instead of having to solve the complete problem of deciding which actions to execute by each individual unit at each time step. One the other hand, large datasets of replays have been created \cite{WeberCig09, synnaeve2012dataset}, from where strategies, trends or plans have been tried to learn. However, StarCraft games are quite complex, and how to automatically learn from such datasets is still an open problem.

\subsubsection{Task Decomposition}
For all the previous reasons, most existing approaches to play games like StarCraft work by decomposing the problem of playing a RTS game into a collection of smaller problems, to be solved independently. Specifically, a common subdivision is:
\begin{itemize}
\item {\em Strategy}: corresponds to the high-level decision making process.  This is
  the  highest level of  abstraction for  the game  comprehension. RTS
  games   are   all   about   finding   an   efficient   strategy   or
  counter-strategy  against a  given opponent.  It concerns  the whole
  set of  units and buildings  one owns. %If  we may allow  ourselves a  military analogy, strategy is orders given by generals at the HQ. 
\item  {\em Tactics}: are  the implementation  of the  current  strategy. It
  implies  army and  building positioning,  movements, timing,  and so
  on.  Tactics concerns a  group of  units. %They  are orders  given by officers on the battlefield.
\item  {\em Reactive  control}: is  the  implementation  of  tactics.  This
  consists   in  moving,   targeting,  firing,   fleeing,  hit-and-run
  techniques (a.k.a. ``kiting'')  during battle. Reactive control focuses
  on a specific unit. %It is soldiers applying officers' orders. 
%\end{itemize}
%
%\noindent
%Both     strategy      and     tactics     are      usually     called
%\textit{macro-management},  whereas reactive  control  is also  called
%\textit{micro-management}. In addition of the three tasks shown above,
%one could also consider these two tasks that an AI has to deal with:
%\begin{itemize}
\item {\em Terrain analysis}: consists  in the analysis of regions composing
  the  map:  choke-points,  minerals  and  gas emplacements,  low  and  high
  walkable grounds, islands, etc.
\item {\em Intelligence gathering}:  corresponds to information collected about
  the  opponent. Because of  the fog-of-war,  players must
  regularly send scouts to localize and spy enemy bases.%, and must see or  infer what buildings and technology the enemy has developed in order  to adapt its strategy.
\end{itemize}

We want to mention also the task decomposition that the gamer community do. Given the fact that players cannot perform multiple actions at the same time, they divide their tasks in two groups:
\begin{itemize}
  \item {\em Macro}: is the ability to control your units individually. A good \emph{macro} player usually keeps their units alive during more time.
  \item {\em Micro}: is the ability to produce units and to expand at the appropriate times to keep your production of units flowing. A good \emph{micro} player usually has the larger army.
\end{itemize}

@@

On the other hand, bots such as Nova or BroodwarBotQ (BBQ) only use a hierarchical organization for {\em micro-management} (controlling the attack units), but use a decentralized organization for the rest of the bot. In Nova and BBQ, there is a collection of modules that control different aspects of the game (workers, production, construction, etc.). These modules can all send actions directly to StarCraft. In Nova those modules coordinate mostly through writing data in a shared blackboard, and in BBQ they coordinate only when they have to use a shared resource (unit) by means of an arbitrator.
